Faster generic IND-CCA secure KEM using encrypt-then-MAC

### 2024-10-13 etm-transform/manuscript/main.tex
More revisions on manuscript

% TODO: describe ML-KEM in appendix
%   Added in Figure 14

% TODO: why is key derived from H(preKey || c') vs H(preKey || t)
% DONE: that was a typo and was fixed. the implementation hashed the tag, not the ciphertext

% TODO: Decapsulation line 9
% DONE: ML-KEM still uses implicit rejection, see line 7 of Algorithm 18 in FIPS 203 (https://nvlpubs.nist.gov/nistpubs/FIPS/NIST.FIPS.203.pdf)

% TODO: Explain the removal of the pseudorandom coin
% DONE: added paragraph in section 3: "Since the encrypt-then-MAC transformation removes re-encryption..."

% TODO: reference for key exchange protocols
% DONE: Added to the beginning of section 5.3

% TODO: revise abstract and introduction
% DONE: revised abstract, reorganized introduction, removed RSA/ElGamal in the last section

% TODO: Why only 128-bit tag?
% DONE: Added a few sentences in section 4.1

@@

%%%% 6. ABSTRACT %%%%
\begin{abstract}
    ML-KEM is a lattice-based IND-CCA secure key encapsulation mechanism (KEM) standardized by NIST in FIPS 203. ML-KEM achieves chosen-ciphertext security by applying the Fujisaki-Okamoto transformation to an IND-CPA secure public-key encryption (PKE) scheme. The Fujisaki-Okamoto transformation uses de-randomization and re-encryption to ensure ciphertext non-malleability, but because ML-KEM's underlying PKE encryption is significantly slower than the underlying PKE decryption, ML-KEM's decapsulation performance is dominated by re-encryption. In this paper, we propose an alternative generic IND-CCA secure KEM transformation that applies the encrypt-then-MAC pattern to an OW-PCA secure PKE and an existentially unforgeable MAC. Compared to the Fujisaki-Okamoto transformation, our encrypt-then-MAC transformation replaces de-randomization and re-encryption with computing a MAC tag. We then implemented $\mlkemplus$ by applying the encrypt-then-MAC transformation to the PKE sub-routines of ML-KEM and a wide selection of MACs including Poly1305, GMAC, CMAC, and KMAC. At all three security levels (128-bit, 196-bit, and 256-bit security), our implementation demonstrates great performance improvements in a variety of practical scenarios. For example, $\mlkemplus$ with Poly1305 reduces decapsulation's CPU cycles by 72\%-79\% and decreases network round trip time in key exchange protocols by 24\%-48\%.
\end{abstract}

@@

\begin{table}[H]
    \centering
    \footnotesize

    \begin{tabular}{|p{1.2cm}|p{1.49cm}|p{1.49cm}|p{1.49cm}|}
        \hline
        & $\text{ML-KEM}^+$ \newline 512 
        & $\text{ML-KEM}^+$ \newline 768 
        & $\text{ML-KEM}^+$ \newline 1024 
        \\
        \hline
        Encap \newline (ccl/tick) 
        & 93157 \newline (+1.8\%) 
        & 146405 \newline (+7.3\%) 
        & 205763 \newline (+3.3\%) 
        \\
        \hline
        Decap \newline (ccl/tick) 
        & 33733 \newline (-72.2\%) 
        & 43315 \newline (-76.8\%) 
        & 51375 \newline (-79.1\%) 
        \\
        \hline
        CT size \newline (bytes) 
        & 784 \newline (+2.1\%) 
        & 1104 \newline (+1.5\%) 
        & 1584 \newline (+1.0\%) 
        \\
        \hline
    \end{tabular}

    \caption{$\mlkemplus$ with Poly1305 achieves large performance improvements in decapsulation at the cost of minimal penalty in encapsulation performance and ciphertext size}\label{tbl:cpu-cycles-summary}
\end{table}

@@

\begin{table}[H]
    \centering
    \footnotesize

    \begin{tabular}{|p{1.6cm}|p{1.49cm}|p{1.49cm}|p{1.49cm}|}
        \hline
        &  $\text{ML-KEM}^+$ \newline 512 
        &  $\text{ML-KEM}^+$ \newline 768 
        &  $\text{ML-KEM}^+$ \newline 1024 
        \\
        \hline
        KE RTT \newline ($\us$) 
        &  70 \newline (-23.9\%) 
        &  99 \newline (-26.7\%) 
        &  138 \newline (-28.5\%) 
        \\
        \hline
        UAKE RTT \newline ($\us$) 
        &  103 \newline (-29.0\%) 
        &  144 \newline (-33.0\%) 
        &  202 \newline (-34.8\%) 
        \\
        \hline
        AKE RTT \newline ($\us$) 
        &  133 \newline (-39.5\%) 
        &  190 \newline (-35.4\%) 
        &  266 \newline (-48.0\%) 
        \\
        \hline
    \end{tabular}

@@

\subsection{Choosing a message authenticator}\label{sec:choosing-a-message-authenticator}
For implementation, we instantiated the MAC with a selection that covered a wide range of MAC designs, including Poly1305 \cite{bernstein2005poly1305}, GMAC \cite{mcgrew2004galois}, CMAC \cite{iwata2003omac}\cite{black2000cbc}, and KMAC \cite{SP80053r4}. All chosen MACs are parameterized with 256-bit keys and 128-bit tags. We chose 256-bit keys for all three security levels (128-bit, 196-bit, and 256-bit security) because choosing a smaller key size does not yield meaningful performance gains (in addition, Poly1305 only supports 256-bit keys). On the other hand, a 128-bit tag size is sufficient for all three security levels because unlike brute-force attacks on the MAC key, which can be carried out offline with large computing capacity, attacks on the tag require interaction with an online peer that plays the role of a MAC verification oracle. This means that the attack bandwidth is limited by the throughput of the online oracle instead of the attacker's computing power, and in practice it is easy to implement rate limiting in the cryptographic protocol such that the adversary cannot submit large quantities of MAC verification queries.

@@

@@

\begin{table}[h]
    \centering
    \footnotesize
    \caption{CPU cycles needed to compute tag on various input sizes}\label{tbl:mac-performance}
    \begin{tabular}{|c|c|c|}
        \hline
        \multicolumn{3}{|c|}{\bf Input size: 768 bytes} \\
        \hline
        MAC & Median & Average \\
        \hline
        Poly1305 & 909 & 2823 \\
        \hline
        GMAC & 3899 & 4859 \\
        \hline
        CMAC & 6291 & 6373 \\
        \hline
        KMAC & 6373 & 7791 \\
        \hline
    \end{tabular}
    \begin{tabular}{|c|c|c|}
        \hline
        \multicolumn{3}{|c|}{\bf Input size: 1088 bytes} \\
        \hline
        MAC & Median & Average \\
        \hline
        Poly1305 & 961 & 2704 \\
        \hline
        GMAC & 3899 & 4827 \\
        \hline
        CMAC & 7305 & 7588 \\
        \hline
        KMAC & 9697 & 9928 \\
        \hline
    \end{tabular}
    \begin{tabular}{|c|c|c|}
        \hline
        \multicolumn{3}{|c|}{\bf Input size: 1568 bytes} \\
        \hline
        MAC & Median & Average \\
        \hline
        Poly1305 & 1065 & 1809 \\
        \hline
        GMAC & 4055 & 5026 \\
        \hline
        CMAC & 8735 & 8772 \\
        \hline
        KMAC & 11647 & 12186 \\
        \hline
    \end{tabular}
\end{table}

@@

    \begin{tabular}[t]{|cc|}
        \hline
        \multicolumn{2}{|c|}{\bf 192-bit security} \\
        \hline
        \multicolumn{2}{|c|}{size parameters (bytes)} \\
        $\pk$ size & 1184 \\
        $\sk$ size & 2400 \\
        \text{ct} size & 1088 \\
        \hline
        \multicolumn{2}{|c|}{$\keygen$ cycles/tick} \\
        Median & 129895 \\
        Average & 130650 \\
        \hline
    \end{tabular}
    \begin{tabular}[t]{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant} 
        & \multicolumn{2}{|c|}{$\encap$ cycles/tick} 
        & \multicolumn{2}{|c|}{$\decap$ cycles/tick} \\
        \cline{2-5}
        & Median & Average & Median & Average \\
        \hline
        \text{ML-KEM-768} & 136405 & 147400 & 186445 & 187529 \\
        \hline
        \text{Kyber768} & 153061 & 153670 & 182129 & 182755 \\
        \hline
        $\text{ML-KEM}^+$-768 w/ Poly1305 & 146405 & 146860 & 43315 & 43463 \\
        \hline
        $\text{ML-KEM}^+$-768 w/ GMAC & 149525 & 150128 & 46513 & 46706 \\
        \hline
        $\text{ML-KEM}^+$-768 w/ CMAC & 153139 & 153735 & 49841 & 50074 \\
        \hline
        $\text{ML-KEM}^+$-768 w/ KMAC & 155219 & 155848 & 52415 & 52611 \\
        \hline
    \end{tabular}\vspace{0.3cm}

    \begin{tabular}[t]{|cc|}
        \hline
        \multicolumn{2}{|c|}{\bf 256-bit security} \\
        \hline
        \multicolumn{2}{|c|}{size parameters (bytes)} \\
        $\pk$ size & 1568 \\
        $\sk$ size & 3168 \\
        \text{ct} size & 1568 \\
        \hline
        \multicolumn{2}{|c|}{$\keygen$ cycles/tick} \\
        Median & 194921 \\
        Average & 195465 \\
        \hline
    \end{tabular}
    \begin{tabular}[t]{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant} 
        & \multicolumn{2}{|c|}{$\encap$ cycles/tick} 
        & \multicolumn{2}{|c|}{$\decap$ cycles/tick} \\
        \cline{2-5}
        & Median & Average & Median & Average \\
        \hline
        \text{ML-KEM-1024} & 199185 & 199903 & 246245 & 247320 \\
        \hline
        \text{Kyber1024} & 222351 & 223260 & 258231 & 259067 \\
        \hline
        $\text{ML-KEM}^+$-1024 w/ Poly1305 & 205763 & 206499 & 51375 & 51562 \\
        \hline
        $\text{ML-KEM}^+$-1024 w/ GMAC & 208805 & 209681 & 54573 & 54780 \\
        \hline
        $\text{ML-KEM}^+$-1024 w/ CMAC & 213667 & 214483 & 59175 & 59408 \\
        \hline
        $\text{ML-KEM}^+$-1024 w/ KMAC & 216761 & 217468 & 62269 & 62516 \\
        \hline
    \end{tabular}
\end{table}

@@

    \begin{tabular}{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant}
        & \multirow{2}{*}{Client TX bytes}
        & \multirow{2}{*}{Server TX bytes}
        & \multicolumn{2}{|c|}{RTT ($\us$)} \\
        \cline{4-5}
        & & & Median & Average \\
        \hline
        \texttt{ML-KEM-512} & 800 & 768 & 92 & 97 \\
        \hline
        $\texttt{ML-KEM-512}^+$ w/ Poly1305 & 800 & 784 & 70 & 72 \\
        \hline
        $\texttt{ML-KEM-512}^+$ w/ GMAC & 800 & 784 & 73 & 76 \\
        \hline
        $\texttt{ML-KEM-512}^+$ w/ CMAC & 800 & 784 & 75 & 79 \\
        \hline
        $\texttt{ML-KEM-512}^+$ w/ KMAC & 800 & 784 & 76 & 78 \\
        \hline
    \end{tabular}\vspace{0.3cm}
    \begin{tabular}{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant}
        & \multirow{2}{*}{Client TX bytes}
        & \multirow{2}{*}{Server TX bytes}
        & \multicolumn{2}{|c|}{RTT ($\us$)} \\
        \cline{4-5}
        & & & Median & Average \\
        \hline
        \texttt{ML-KEM-768} & 1184 & 1088 & 135 & 140 \\
        \hline
        $\texttt{ML-KEM-768}^+$ w/ Poly1305 & 1184 & 1104 & 99 & 104 \\
        \hline
        $\texttt{ML-KEM-768}^+$ w/ GMAC & 1184 & 1104 & 101 & 105 \\
        \hline
        $\texttt{ML-KEM-768}^+$ w/ CMAC & 1184 & 1104 & 103 & 109 \\
        \hline
        $\texttt{ML-KEM-768}^+$ w/ KMAC & 1184 & 1104 & 103 & 107 \\
        \hline
    \end{tabular}\vspace{0.3cm}
    \begin{tabular}{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant}
        & \multirow{2}{*}{Client TX bytes}
        & \multirow{2}{*}{Server TX bytes}
        & \multicolumn{2}{|c|}{RTT ($\us$)} \\
        \cline{4-5}
        & & & Median & Average \\
        \hline
        \texttt{ML-KEM-1024} & 1568 & 1568 & 193 & 199 \\
        \hline
        $\texttt{ML-KEM-1024}^+$ w/ Poly1305 & 1568 & 1584 & 138 & 141 \\
        \hline
        $\texttt{ML-KEM-1024}^+$ w/ GMAC & 1568 & 1584 & 140 & 145 \\
        \hline
        $\texttt{ML-KEM-1024}^+$ w/ CMAC & 1568 & 1584 & 143 & 148 \\
        \hline
        $\texttt{ML-KEM-1024}^+$ w/ KMAC & 1568 & 1584 & 144 & 149 \\
        \hline
    \end{tabular}
\end{table}

>>>>

% TODO: add the implementation of the protocols in the abstract.
% DONE: added "We also implemented key exchange protocols and measured..."

% TODO: remove 768/1024 security level data, add remarks about 128-bit tag security.
% DONE: removed Poly1305 data with ML-KEM-768/1024
%   GCM's tag is up to 128 bits: https://nvlpubs.nist.gov/nistpubs/legacy/sp/nistspecialpublication800-38d.pdf
%   CMAC's tag is up to 128 bits because it uses AES as underlying block cipher

@@

%%%% 6. ABSTRACT %%%%
\begin{abstract}
    ML-KEM is a lattice-based IND-CCA secure key encapsulation mechanism (KEM) standardized by NIST in FIPS 203. ML-KEM achieves chosen-ciphertext security by applying the Fujisaki-Okamoto transformation to an IND-CPA secure public-key encryption (PKE) scheme. The Fujisaki-Okamoto transformation uses de-randomization and re-encryption to ensure ciphertext non-malleability, but because ML-KEM's underlying PKE encryption is significantly slower than the underlying PKE decryption, ML-KEM's decapsulation performance is dominated by re-encryption. In this paper, we propose an alternative generic IND-CCA secure KEM transformation that applies the encrypt-then-MAC pattern to an OW-PCA secure PKE and an existentially unforgeable MAC. Compared to the Fujisaki-Okamoto transformation, our encrypt-then-MAC transformation replaces de-randomization and re-encryption with computing a MAC tag. We then implemented $\mlkemplus$ by applying the encrypt-then-MAC transformation to the PKE sub-routines of ML-KEM and a wide selection of MACs including Poly1305, GMAC, CMAC, and KMAC. At the cost of minimal increase in encapsulation CPU cycles (+1.8\%) and ciphertext size (+2.1\%), $\mlkemplus$ achieves massive reduction of decapsulation CPU cycles (-72.2\%) compared to ML-KEM. We also implemented key exchange protocols and measured realistic network round trip time (RTT), where $\mlkemplus$ reduces RTT by 23.9\%-39.5\% compared to ML-KEM.
\end{abstract}

@@

\begin{table}[H]
    \centering
    % \footnotesize
    \small

    % TODO: Poly1305 may not be suitable for 192/256 bit security level
    % \begin{tabular}{|p{1.2cm}|p{1.49cm}|p{1.49cm}|p{1.49cm}|}
    %     \hline
    %     & $\text{ML-KEM}^+$ \newline 512 
    %     & $\text{ML-KEM}^+$ \newline 768 
    %     & $\text{ML-KEM}^+$ \newline 1024 
    %     \\
    %     \hline
    %     Encap \newline (ccl/tick) 
    %     & 93157 \newline (+1.8\%) 
    %     & 146405 \newline (+7.3\%) 
    %     & 205763 \newline (+3.3\%) 
    %     \\
    %     \hline
    %     Decap \newline (ccl/tick) 
    %     & 33733 \newline (-72.2\%) 
    %     & 43315 \newline (-76.8\%) 
    %     & 51375 \newline (-79.1\%) 
    %     \\
    %     \hline
    %     CT size \newline (bytes) 
    %     & 784 \newline (+2.1\%) 
    %     & 1104 \newline (+1.5\%) 
    %     & 1584 \newline (+1.0\%) 
    %     \\
    %     \hline
    % \end{tabular}
    \begin{tabular}{|p{2.5cm}|p{1.4cm}|p{1.4cm}|p{1.4cm}|}
       \hline
       & Encap \newline (ccl/tick) 
       & Decap \newline (ccl/tick) 
       & CT size \newline (bytes) \\
       \hline
       $\mlkemplus$-512 \newline (Poly1305)
       & 93157 \newline (+1.8\%)
       & 33733 \newline (-72.2\%)
       & 784 \newline (+2.1\%) \\
       \hline
       $\mlkemplus$-512 \newline (GMAC)
       & 97369 \newline (+6.5\%)
       & 37725 \newline (-68.9\%)
       & 784 \newline (+2.1\%) \\
       \hline
       $\mlkemplus$-512 \newline (CMAC)
       & 99739 \newline (+9.0\%)
       & 40117 \newline (-66.9\%)
       & 784 \newline (+2.1\%) \\
       \hline
       $\mlkemplus$-512 \newline (KMAC-256)
       & 101009 \newline (+10.4\%)
       & 40741 \newline (-66.4\%)
       & 784 \newline (+2.1\%) \\
       \hline
    \end{tabular}

    \caption{$\mlkemplus$ achieves large performance improvements in decapsulation at the cost of minimal penalty in encapsulation performance and ciphertext size}\label{tbl:cpu-cycles-summary}
\end{table}

@@

\begin{table}[H]
    \centering
    % \footnotesize
    \small

    % TODO: Poly1305 might not be suitable for 192/256 bit security
    % \begin{tabular}{|p{1.6cm}|p{1.49cm}|p{1.49cm}|p{1.49cm}|}
    %     \hline
    %     &  $\text{ML-KEM}^+$ \newline 512 
    %     &  $\text{ML-KEM}^+$ \newline 768 
    %     &  $\text{ML-KEM}^+$ \newline 1024 
    %     \\
    %     \hline
    %     KE RTT \newline ($\us$) 
    %     &  70 \newline (-23.9\%) 
    %     &  99 \newline (-26.7\%) 
    %     &  138 \newline (-28.5\%) 
    %     \\
    %     \hline
    %     UAKE RTT \newline ($\us$) 
    %     &  103 \newline (-29.0\%) 
    %     &  144 \newline (-33.0\%) 
    %     &  202 \newline (-34.8\%) 
    %     \\
    %     \hline
    %     AKE RTT \newline ($\us$) 
    %     &  133 \newline (-39.5\%) 
    %     &  190 \newline (-35.4\%) 
    %     &  266 \newline (-48.0\%) 
    %     \\
    %     \hline
    % \end{tabular}
    \begin{tabular}{|p{2.5cm}|p{1.4cm}|p{1.8cm}|p{1.8cm}|}
       \hline
       & KE RTT \newline ($\us$) 
       & UAKE RTT \newline ($\us$) 
       & AKE RTT \newline ($\us$) \\
       \hline
       $\mlkemplus$-512 \newline (Poly1305)
       & 70 \newline (-23.9\%)
       & 103 \newline (-29.0\%)
       & 133 \newline (-39.5\%) \\
       \hline
       $\mlkemplus$-512 \newline (GMAC)
       & 73 \newline (-20.7\%)
       & 106 \newline (-26.9\%)
       & 139 \newline (-36.8\%) \\
       \hline
       $\mlkemplus$-512 \newline (CMAC)
       & 75 \newline (-18.5\%)
       & 108 \newline (-25.5\%)
       & 143 \newline (-35.0\%) \\
       \hline
       $\mlkemplus$-512 \newline (KMAC-256)
       & 76 \newline (-17.4\%)
       & 109 \newline (-24.8\%)
       & 145 \newline (-34.1\%) \\
       \hline
    \end{tabular}

@@

\subsection{Choosing a message authenticator}\label{sec:choosing-a-message-authenticator}
For implementation, we instantiated the MAC with a selection that covered a wide range of MAC designs, including Poly1305 \cite{bernstein2005poly1305}, GMAC \cite{mcgrew2004galois}, CMAC \cite{iwata2003omac}\cite{black2000cbc}, and KMAC \cite{SP80053r4}.

% All chosen MACs are parameterized with 256-bit keys and 128-bit tags. We chose 256-bit keys for all three security levels (128-bit, 196-bit, and 256-bit security) because choosing a smaller key size does not yield meaningful performance gains (in addition, Poly1305 only supports 256-bit keys). On the other hand, a 128-bit tag size is sufficient for all three security levels because unlike brute-force attacks on the MAC key, which can be carried out offline with large computing capacity, attacks on the tag require interaction with an online peer that plays the role of a MAC verification oracle. This means that the attack bandwidth is limited by the throughput of the online oracle instead of the attacker's computing power, and in practice it is easy to implement rate limiting in the cryptographic protocol such that the adversary cannot submit large quantities of MAC verification queries.

@@

\emph{Remark.} Unfortunately Poly1305, GMAC, and CMAC all use tag size of 128 bits with no straightforward ways to increase the tag size. We suspect this small tag size to be unsuitable for instantiation with ML-KEM-768 and ML-KEM-1024 due to their higher security target (192-bit and 256-bit security respectively). Consequently, we only instantiate with ML-KEM-512, with a target 128-bit security level.

@@

\begin{table}[h]
    \centering
    % \footnotesize
    \small
    \caption{CPU cycles needed to compute tag on various input sizes}\label{tbl:mac-performance}
    \begin{tabular}{|c|c|c|}
        \hline
        \multicolumn{3}{|c|}{\bf Input size: 768 bytes} \\
        \hline
        MAC & Median (ccl/tick) & Average (ccl/tick)\\
        \hline
        Poly1305 & 909 & 2823 \\
        \hline
        GMAC & 3899 & 4859 \\
        \hline
        CMAC & 6291 & 6373 \\
        \hline
        KMAC & 6373 & 7791 \\
        \hline
    \end{tabular}
    % TODO: 128-bit tag size is not suitable for higher security level.
    % \begin{tabular}{|c|c|c|}
    %     \hline
    %     \multicolumn{3}{|c|}{\bf Input size: 1088 bytes} \\
    %     \hline
    %     MAC & Median & Average \\
    %     \hline
    %     Poly1305 & 961 & 2704 \\
    %     \hline
    %     GMAC & 3899 & 4827 \\
    %     \hline
    %     CMAC & 7305 & 7588 \\
    %     \hline
    %     KMAC & 9697 & 9928 \\
    %     \hline
    % \end{tabular}
    % \begin{tabular}{|c|c|c|}
    %     \hline
    %     \multicolumn{3}{|c|}{\bf Input size: 1568 bytes} \\
    %     \hline
    %     MAC & Median & Average \\
    %     \hline
    %     Poly1305 & 1065 & 1809 \\
    %     \hline
    %     GMAC & 4055 & 5026 \\
    %     \hline
    %     CMAC & 8735 & 8772 \\
    %     \hline
    %     KMAC & 11647 & 12186 \\
    %     \hline
    % \end{tabular}
\end{table}

@@

    % TODO: 128-bit tag size is unsuitable for higher security level
    % \begin{tabular}[t]{|cc|}
    %     \hline
    %     \multicolumn{2}{|c|}{\bf 192-bit security} \\
    %     \hline
    %     \multicolumn{2}{|c|}{size parameters (bytes)} \\
    %     $\pk$ size & 1184 \\
    %     $\sk$ size & 2400 \\
    %     \text{ct} size & 1088 \\
    %     \hline
    %     \multicolumn{2}{|c|}{$\keygen$ cycles/tick} \\
    %     Median & 129895 \\
    %     Average & 130650 \\
    %     \hline
    % \end{tabular}
    % \begin{tabular}[t]{|c|c|c|c|c|}
    %     \hline
    %     \multirow{2}{*}{KEM variant} 
    %     & \multicolumn{2}{|c|}{$\encap$ cycles/tick} 
    %     & \multicolumn{2}{|c|}{$\decap$ cycles/tick} \\
    %     \cline{2-5}
    %     & Median & Average & Median & Average \\
    %     \hline
    %     \text{ML-KEM-768} & 136405 & 147400 & 186445 & 187529 \\
    %     \hline
    %     \text{Kyber768} & 153061 & 153670 & 182129 & 182755 \\
    %     \hline
    %     $\text{ML-KEM}^+$-768 w/ Poly1305 & 146405 & 146860 & 43315 & 43463 \\
    %     \hline
    %     $\text{ML-KEM}^+$-768 w/ GMAC & 149525 & 150128 & 46513 & 46706 \\
    %     \hline
    %     $\text{ML-KEM}^+$-768 w/ CMAC & 153139 & 153735 & 49841 & 50074 \\
    %     \hline
    %     $\text{ML-KEM}^+$-768 w/ KMAC & 155219 & 155848 & 52415 & 52611 \\
    %     \hline
    % \end{tabular}\vspace{0.3cm}

    % \begin{tabular}[t]{|cc|}
    %     \hline
    %     \multicolumn{2}{|c|}{\bf 256-bit security} \\
    %     \hline
    %     \multicolumn{2}{|c|}{size parameters (bytes)} \\
    %     $\pk$ size & 1568 \\
    %     $\sk$ size & 3168 \\
    %     \text{ct} size & 1568 \\
    %     \hline
    %     \multicolumn{2}{|c|}{$\keygen$ cycles/tick} \\
    %     Median & 194921 \\
    %     Average & 195465 \\
    %     \hline
    % \end{tabular}
    % \begin{tabular}[t]{|c|c|c|c|c|}
    %     \hline
    %     \multirow{2}{*}{KEM variant} 
    %     & \multicolumn{2}{|c|}{$\encap$ cycles/tick} 
    %     & \multicolumn{2}{|c|}{$\decap$ cycles/tick} \\
    %     \cline{2-5}
    %     & Median & Average & Median & Average \\
    %     \hline
    %     \text{ML-KEM-1024} & 199185 & 199903 & 246245 & 247320 \\
    %     \hline
    %     \text{Kyber1024} & 222351 & 223260 & 258231 & 259067 \\
    %     \hline
    %     $\text{ML-KEM}^+$-1024 w/ Poly1305 & 205763 & 206499 & 51375 & 51562 \\
    %     \hline
    %     $\text{ML-KEM}^+$-1024 w/ GMAC & 208805 & 209681 & 54573 & 54780 \\
    %     \hline
    %     $\text{ML-KEM}^+$-1024 w/ CMAC & 213667 & 214483 & 59175 & 59408 \\
    %     \hline
    %     $\text{ML-KEM}^+$-1024 w/ KMAC & 216761 & 217468 & 62269 & 62516 \\
    %     \hline
    % \end{tabular}
\end{table}

@@

    \begin{tabular}{|c|c|c|c|c|}
        \hline
        \multirow{2}{*}{KEM variant}
        & \multirow{2}{*}{Client TX bytes}
        & \multirow{2}{*}{Server TX bytes}
        & \multicolumn{2}{|c|}{RTT ($\us$)} \\
        \cline{4-5}
        & & & Median & Average \\
        \hline
        ML-KEM-512 & 800 & 768 & 92 & 97 \\
        \hline
        $\mlkemplus$-512 w/ Poly1305 & 800 & 784 & 70 & 72 \\
        \hline
        $\mlkemplus$-512 w/ GMAC & 800 & 784 & 73 & 76 \\
        \hline
        $\mlkemplus$-512 w/ CMAC & 800 & 784 & 75 & 79 \\
        \hline
        $\mlkemplus$-512 w/ KMAC & 800 & 784 & 76 & 78 \\
        \hline
    \end{tabular}\vspace{0.3cm}
    % TODO: 128-bit tag size is unsuitable for higher security level
    % \begin{tabular}{|c|c|c|c|c|}
    %     \hline
    %     \multirow{2}{*}{KEM variant}
    %     & \multirow{2}{*}{Client TX bytes}
    %     & \multirow{2}{*}{Server TX bytes}
    %     & \multicolumn{2}{|c|}{RTT ($\us$)} \\
    %     \cline{4-5}
    %     & & & Median & Average \\
    %     \hline
    %     \texttt{ML-KEM-768} & 1184 & 1088 & 135 & 140 \\
    %     \hline
    %     $\texttt{ML-KEM-768}^+$ w/ Poly1305 & 1184 & 1104 & 99 & 104 \\
    %     \hline
    %     $\texttt{ML-KEM-768}^+$ w/ GMAC & 1184 & 1104 & 101 & 105 \\
    %     \hline
    %     $\texttt{ML-KEM-768}^+$ w/ CMAC & 1184 & 1104 & 103 & 109 \\
    %     \hline
    %     $\texttt{ML-KEM-768}^+$ w/ KMAC & 1184 & 1104 & 103 & 107 \\
    %     \hline
    % \end{tabular}\vspace{0.3cm}
    % \begin{tabular}{|c|c|c|c|c|}
    %     \hline
    %     \multirow{2}{*}{KEM variant}
    %     & \multirow{2}{*}{Client TX bytes}
    %     & \multirow{2}{*}{Server TX bytes}
    %     & \multicolumn{2}{|c|}{RTT ($\us$)} \\
    %     \cline{4-5}
    %     & & & Median & Average \\
    %     \hline
    %     \texttt{ML-KEM-1024} & 1568 & 1568 & 193 & 199 \\
    %     \hline
    %     $\texttt{ML-KEM-1024}^+$ w/ Poly1305 & 1568 & 1584 & 138 & 141 \\
    %     \hline
    %     $\texttt{ML-KEM-1024}^+$ w/ GMAC & 1568 & 1584 & 140 & 145 \\
    %     \hline
    %     $\texttt{ML-KEM-1024}^+$ w/ CMAC & 1568 & 1584 & 143 & 148 \\
    %     \hline
    %     $\texttt{ML-KEM-1024}^+$ w/ KMAC & 1568 & 1584 & 144 & 149 \\
    %     \hline
    % \end{tabular}
\end{table}
